---
title: "The hidden manifold distance for functional data"
output:
  html_document:
    df_print: paged
    number_sections: yes
  word_document: default
  pdf_document:
    keep_tex: yes
    number_sections: yes
bibliography: references.bib
---

\newcommand {\To}{\rightarrow}
\newcommand {\TO}{\Rightarrow}
\newcommand {\R}{\mathbb{R}}
\newcommand {\Prob}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand {\cov}{\textrm{Cov}}
\newcommand {\var}{\textrm{Var}}
\newcommand {\1}{\textrm{\textbf{1}}}
\newcommand{\M}{\mathcal{M}}

# Abstract
Functional data analysis is the statistical analysis of smooth infinite-dimensional curves. 
The challenge of analyzing infinite-dimensional objects is ameliorated by the fact that the dimensionality of curves is only artifically high if they are smooth.
There is also another sense in which the dimensionality of the curves under study may be low. This is the much less explored idea that curves may actually live in a submanifold with low intrinsic dimension.
Interesting classes of functional manifold data include classes of probability density functions and classes of warped curves of a common template function. 
Since these spaces are not vector spaces, even basic operations such as addition and subtraction require special consideration. 
In this work we set ourselves the task of estimation the pairwise geodesic distance between functional manifold data when we only have access to their noisy realizations which live near a non-linear manifold rather than exactly on the manifold. 
<!-- This setting cannot be tackled by classic manifold learning techniques which require data to live exactly on a manifold.  -->
<!-- We will show that estimation of the geodesic distance cannot easily be accomplished by a naive method such as smoothing followed by a shortest-path algorithm. -->
The proposed methodology first projects the functional data on to the underlying hidden manifold and then performs operations on this hidden manifold. 
Good estimation of the pairwise geodesic distance has beneficial implications for many downstream tasks in functional data analysis which we illustrate in the case of distanced-based functional classification.


# Introduction 
<!-- Susan  -->
Many statistical methods depend on a measure of distance. Methods in functional data analysis, the study of infinite-dimensional smooth curves, is no exception. In situations where the manifold hypothesis might be plausible, one might consider using a geodesic distance that takes into account the intrinsic structure of the manifold rather than the standard $L^2$ distance. The benefits of adopting the geodesic distance may be realized in downstream tasks such as distanced-based clustering and classification of functional obvervations.

Specifically, let $X_1,\ldots,X_n$ be a sample of $n$ independant realizations of a random variable $X$ that takes value in the Hilbert space $L^2([a,b],\R)$. Suppose additionally that the function $X$ belongs to a low-dimensional nonlinear Riemannian manifold $\M \subset L^2([a,b],\R)$ where $g$ is a Riemannian metric tensor on $\M$ which can be used to assign a metric on the manifold as follows [@Lin2014]. For each point $X$ on the manifold, the Riemannian metric tensor $g$ has an inner product $g_X$ on the tangent space $T_X \M$. The norm of a tangent vector $V \in T_X \M$ is defined as $$||V|| = \sqrt{g_X(V,V)}.$$ The geodesic distance between two functions $X_i,X_j$ on the manifold $\M$, based on this metric tensor $g$, is defined as
$$ d_g(X_i,X_j):=\inf \{l(\gamma): \gamma:[a,b] \to \M \text{ piecewise smooth}, \gamma(a) = X_i, \gamma(b) = X_j]  \}$$
where $$ l(\gamma) := \int_a^b || \frac{\,d\gamma}{\,d t}(t) || \,dt$$
is the length of a smooth curve $\gamma: [a,b] \subset \R \to \M$.



<!-- ???insert example that shows pairwise geodesic distances give visually more interesting results than pairwise $L_2$ distance.??? -->

<!-- (gives examples and references for functional manifolds). -->
<!-- Before continuing, some diambiguities are called for. Contrast existing work on functional data where either the domain is a manifold or the range is a manifold, or both.  -->

<!-- There exist many nonlinear dimension reduction methods for manifolds embedded in Euclidean space, also known as manifold learning methods, which are particularly popular in computer vision. However, the success of these methods typically require the data to be observed with a high signal-to-noise ratio. (is it true? ref?)  -->

Given perfect functional observations $X_i$ and $X_j$, i.e.\ with no measurement error and on a very dense domain grid, the estimation of the geodesic distance $d_g(X_i,X_j)$ can be accomplished straightforwardly using a shortest-path algorithm, e.g. the Floyd-Warhsall algorithm [@Floyd1962]. 
However, when functional data are observed with noise, this approach is likely to fail. This is because the error may be such that the noisy functional manifold data no longer live on a manifold. 
A preprocessing step where the nosiy functional manifold data is smoothed is also likely to fail. This is because smoothing techniques in functional data analysis are designed for the $L_2$ recovery of the original curve, not to ensure the recovered functional versions of the data lie on or close to the functional manifold $\M$. 
<!-- This presents challenges to the application of many manifold learning techniques which assume fully observed nearly-noiseless inputs. -->

<!-- Estimating the manifold $\M$ from a sample of functions observed discretely and with measurements is a challenging task (results depend on curvature of the manifold, need a lot of data). Even after recovering a functional version of the data, the chances are pretty high that $\tilde X_1,\ldots,\tilde X_n$ do not lie exactly on $\M$.  -->

In this work, we put forth a technique for the specific task of recovering pairwise geodesic distances $\{ d_\M (X_i,X_j)\}_{i>j}$ when we have, in lieu of $X_i$ and $X_j$, access only to discretely-observed noisy functional observations $Y_i$ and $Y_j$ that possibly live off the true manifold. This is accomplished by first projecting the functional data on to the underlying hidden manifold and then performing operations on this hidden manifold. 

<!-- Pairwise distances are important for many downstream tasks, e.g.  classification, clustering, manifold learning (ISOMAP), nonparametric regression (kernel regression). Certainly, pairwise geodesic distances are legitimate objects of interest in themselves but our motivation is more general in nature. Specifically, we use pairwise distances as an illuminating example to advocate and promote the nonlinear perspective in FDA.  -->


## Related Work
<!-- Susan  -->
[@ChenMuller2012]

[@LinYao2017]

[@Dimeglio2014]

# Proposed method for estimating geodesic distances [Marie]

Suppose that each curve $X_i$ is observed with measurements errors on a grid $T_i=(t_{i1},\ldots,t_{iK})$, i.e. we observe a sample of $K$-dimensional vectors $Y_1,\ldots,Y_n$ with $Y_{ij} = X_i(t_{ij}) + \epsilon_{ij}$, where the random variables $\epsilon_{ij}$ are of mean zero and uncorrelated with each other. We assume that the grids $T_1,\ldots,T_n$ are dense.

Let $\tilde X_1,\ldots,\tilde X_n$ denote the functional versions of the raw data by some smoothing method. Our proposed methodology is unspecific to the smoothing method employed. For simplicity however, let's say we employ spline smoothing to recover the functional versions of the raw data, i.e.\
$$ \tilde X_i = \arg\min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\}$$
where $\lambda>0$ is a tuning parameter controlling the smoothness of $\tilde X_i$. 


We first describe an important helper algorithm which we call IsoGeo. It is a subroutine of the classic Isomap procedure.
Isomap is a three steps procedure that takes as input a set of points $x_1,\ldots,x_n\in \R^D$ and produces an embedding of the input data in the space $\R^d$ with $d<D$, that preserves pairwise geodesic distances. 

\begin{enumerate}
\item The first step consists in constructing a weighted graph $G$ with nodes corresponding to the input data and for which the weight of an edge between two nodes $x_i$ and $x_j$ is equal to $||x_i-x_j ||_{\R^D}$. The graph $G$ is constructed such that a node is connected to a fix number $N$ of neighbors or to all nodes that are at a distance smaller than a given value $\epsilon$.
\item In the second step, the pairwise geodesic distances $d_\M(x_i,x_j)$
are estimated based on $G$. The geodesic distance between two nodes is simply the length of the shortest path in $G$ between these two nodes, which we calculate with the Floyd-Warshall algorithm [@Floyd1962].
<!-- Which one do we actually use in our implementation, Floyd-Warshall or Dijkstra's? -->
\item The last step consists in using MDS on the matrix of pairwise geodesic distances obtained in the first step in order to obtain an embedding in $\R^d$.
\end{enumerate}

In what follows, we call IsoGeo the procedure of estimating pairwise geodesic distances with the two first steps of Isomap. Note that Floyd or Dijkstra are methods to find the smallest path given a weigthed graph but they are not related to how to calculate that graph (which is the difficult part, depend on number of neighbors, many method out there to calculate that graph in a robust way, P-ISOMAP is one of these). 

Our method is described by the following steps:

<!-- 1. [Susan: I think the smoothing step should be independent of our proposed methodology] Transform each vector $Y_i$ into a function $\tilde X_i$ by spline smoothing: -->
<!-- $$ \tilde X_i = \arg\min_{f\in C^2[0,1]}\left\{\sum_{j=1}^{K}\left(f(t_{ij})-Y_{ij}\right)^2+\lambda \|\partial^2_tf\|^2_{L^2}\right\}$$ -->
<!-- where $\lambda>0$ is a tuning parameter controlling the smoothness of $\tilde X_i$.  -->
2. Obtain a $s$-dimensional representation $\tilde X^s_1,\ldots,\tilde X^s_n$ of the functions $\tilde X_1,\ldots,\tilde X_n$ that "preserves" the pairwise $L^2$ distances by using MDS.
3. Obtain $\tilde X^{s,\hat \M}_1,\ldots,\tilde X^{s,\hat \M}_n$ a "projection" of $\tilde X^s_1,\ldots,\tilde X^s_n$ onto a ridge $\hat \M$ which is computed with the subspace constrained mean-shift algorithm [@Ozertem2011].
4. Use  IsoGeo to approximate the pairwise geodesic distances $\{d_{\hat \M}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j)\}_{i>j}$ and define the $n$ by $n$ matrix 
$$
\hat G(i,j)=\hat G(j,i) = \left\{ \begin{array}{ll}
 d_{\hat \M}(\tilde X^{s,\hat \M}_i,\tilde X^{s,\hat \M}_j) & \textrm{if $i\neq j$,}\\
 0 & \textrm{otherwise.}
  \end{array} \right.
$$


Since the ridge estimation obtained from noisy measurements of a manifold should approximate well the manifold [@Genovese2014], we expect the points $\tilde X^{s,\hat \M}_1,\ldots,\tilde X^{s,\hat \M}_n$ to lie close to the real manifold $\M$ and then $d_{\hat \M}$ to be close to $d_\M$. Ridge estimation suffers from the curse of dimensionality, this is why we first reduce the dimension of our data with MDS and then apply the shift-mean algorithm to estimate the ridge.


## Selection of tuning parameters [Marie]
Describe heuristics. We pick the bandwidth $h$ in the subspace constrained mean-shift using Equation (A1) of [@ChenHo2015].

<!-- ##############  -->

# Simulation study

We perform a simulation study to ascertain the efficacy of our method for estimating pairwise geodesic distances when one only has access to discretely-observed noisy functional data. Three different metrics are used to assess the quality of a pairwise geodesic distance estimator. 

* Near-$\epsilon$ isometry: area under the curve of the ROC curve with $\epsilon$ on the $x$-axis and degree to which near-$\epsilon$ isometry holds, i.e.\ the percentage of estimated pairwise distances between $1-\epsilon$ and $1+\epsilon$ of the truth pairwise distance.  
* Relative Frobenius: let $d$ and $\hat d$ be the true and estimated $n \times n$ geodesic distance matrices, respectively. 
$$
||d-\hat d||/||d||
$$
* Pearson correlation: take the upper diagonal of $d$ and $\hat d$, calculate their Pearson correlation coefficient. 

## Alternative estimators of geodesic distance [Marie]

* **(Raw Data) RD** Apply IsoGeo on the raw data $Y_1,\ldots,Y_n$ to obtain an estimator $\hat G_{\textrm{RD}}$. Note that this procedure can only be used if the grid $T_i$ is the same for each $i$.
* **(Spline Smoothing) SS** Transform each vector $Y_i$ into a function $\tilde X_i$ by spline smoothing. Apply IsoGeo on the vectors $\{(\tilde X_i(t_1), \ldots, X_i(t_K)\}_{i=1}^n$, where $t_1,\ldots,t_K$ is a regular grid of $[a,b]$, to obtain an estimator $\hat G_{\textrm{SS}}$.
* **(P-ISOMAP) pI** Transform each vector $Y_i$ into a function $\tilde X_i$ by spline smoothing. Calculate the weighted graph of step I of Isomap with a penalty (describe the penalty of Chen and Muller). Apply step II of Isomap to obtain $\hat G_{\textrm{pI}}$.
* **(Random Projection) RP**  same method as our but change step 2 by : obtain $s$-dimensional representation by random projection and setp 4 by obtain $\hat G$ using a ensemble method.

We ran Chen and Muller's P-ISOMAP separately in Matlab with their automatic penalty selection and discovered that ???.



## Simulation scenarios 
<!-- Susan  -->

### Isometric functional manifold of normal density functions
<!-- Scenario 5 -->
This scenario is modified from what is referred to as Manifold 2 in [@ChenMuller2012] by fixing the variance of the normal density to be $1$. We have $$\M =  \left \{X_\beta: \beta \in [-1,1], t \in [a,b]\right \}$$ with the $L_2$ inner product as the metric tensor of $\M$, where $X_\beta: [a,b] \to \R$ is given by $X_\beta(t) = \frac{1}{\sqrt{2\pi}} \exp{[-\frac{1}{2}(t-\beta)^2]}$. We set $a=-4$ and $b=4$.The geodesic distance between the curves $X_{\beta_1}$ and $X_{\beta_2}$ is given by
\begin{eqnarray*}
d(X_{\beta_1},X_{\beta_2}) &=& \int_{\beta_1}^{\beta_2} \left \| \frac{d X_\beta (t)}{d\beta} \right\|_{L^2} d\beta \\
&=&  \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4 \frac{1}{\sqrt{\pi}} \exp\{-(t-\beta)^2\}(t-\beta)^2 dt  }  \  d\beta \\
&=&   \int_{\beta_1}^{\beta_2} \sqrt{ \frac{1}{2\sqrt{\pi}} \int_{-4}^4(t-\beta)^2 f(t) dt  }  \  d\beta, \textrm{ where $f$ is the density of a N$(\beta,1/2)$ }   \\
&\approx&  \int_{\beta_1}^{\beta_2}  \sqrt{ \frac{1}{2\sqrt{\pi}} \frac{1}{2}} \ d\beta \\
&=& (\beta_2-\beta_1) \frac{1}{2\pi^{1/4}},
\end{eqnarray*}
where the approximation comes from the fact that we are integrating on $[a,b]=[-4,4]$ and not on $\R$. 
We can see this manifold is isometric, since the geodesic distance between $X_{\beta_1}$ and $X_{\beta_2}$ in $\M$ is the Euclidan distance between the $\beta$'s, up to some scaling factor. Note that the "straight" line connecting $X_{\beta_1}$ and $X_{\beta_2}$ in $\M$ does not always stay inside of $\M$, so we cannot employ the calculation technique of Scenario 1. 

### Functional manifold of square root velocity functions
It was shown in [@Joshi2007] that the square root representation of probability density functions has a nice closed form geodesic.
They consider the manifold $$ \M = \{ \psi:[0,1] \to \R : \psi \ge 0, \int_0^1 \psi^2(s) \,ds = 1 \}$$ with the metric tensor given by the Fisher-Rao metric tensor $$ <v_1,v_2> = \int_0^1 v_1(s) v_2(s) \,ds $$ for two tangent vectors $v_1,v_2 \in T_\psi(\M)$. Note that this concides with the $L_2[0,1]$ inner product. [@Joshi2007] showed that the geodesic distance between any two $\psi_1$ and $\psi_2$ in $\M$ is simply $$d(\psi_1,\psi_2) = \cos^{-1}<\psi_1,\psi_2>.$$
We will specifically examine the square root of $Beta(\alpha,\beta)$ distributions which is supported on $[0,1]$. That is, $$ M = \{ \psi_{\alpha,\beta}: 1 \le \alpha \le 5, 2 \le \beta \le 5\} $$ where $\psi_{\alpha,\beta}: [0,1] \to \R$ is the pdf of $Beta(\alpha,\beta)$.
 
### Functional manifold of warping functions
<!-- Scenario 5 -->
This is based on Equations (17) and (18) of [@Kneip2008] but with $z_{i1}, z_{i2}$ set to $1$. (Equation 17 has a typo where the exponentials are missing negative signs).Let $X_\alpha(t) = \mu(h_\alpha(t))$ be defined on $[-3,3]$ where $$ \mu(t) = \exp\{(t-1.5)^2/2\} + \exp\{(t+1.5)^2/2\}$$ and $$ h_\alpha(t) = 6 \frac{ \exp\{\alpha(t+3)/6\} - 1}{\exp\{\alpha\}-1}, \alpha \ne 0 $$ and $h_\alpha(t) = t$ if $\alpha = 0$. Consider the manifold $$ M = \{X_\alpha: -1 \le \alpha \le 1\} $$. The geodesic distance is then $$ d(X_{\alpha_1},X_{\alpha_2}) = \int_{\alpha_1}^{\alpha_2} \left \| \frac{d X_\alpha (t)}{d\alpha} \right\|_{L^2} d\alpha.$$

## Sampling from a functional manifold
As of yet, there is little work as to how to sample from a functional manifold. Even in the Euclidean case, it is not obvious how sampling should be done [@Diaconis2013]. To safeguard against sampling unevenly on the functional manifold, we sample on a very concentrated measure for the intrinsic parameters.
For both the manifold of normal densities and the manifold of warping functions, we sample $\alpha$ according to ???rtruncnorm(samplesize,alphamin,alphamax,(alphamin+alphamax)/2,1)???
For the manifold of square root beta densities, we sample as follows       
alpha<- rtruncnorm(nb_alpha,alphamin,alphamax,(alphamax+alphamin)/2,0.3)
beta<- rtruncnorm(nb_beta,betamin,betamax,(betamax+betamin)/2,0.3)
      
How to sample properly from a functional manifold could be interesting future work.


## Results
Submit compare_methods.sh as slurm job using sbatch.
```{r, engine = 'bash', eval = FALSE}
sbatch compare_methods.sh
```

Alternatively, we can run the R script compare_methods.R locally by removing the slurm_arrayid chunk
```{r, eval=FALSE}
R compare_methods.R
```

Simulation results are then visualized by invoking cluster_results.R. There's some unfortunate hardcoding still in cluster_results.R
```{r, eval=FALSE}
R cluster_results.R
```
<!--

## The cost of employing proposed method when manifold is flat [Susan/Marie?]

Suppose we take scenario 1 where the geodesic distance coincides with the $L_2$ distance. Then what is the cost of employing our distance estimator compared to smoothing the data and doing numerical integration to find the $L_2$ distance?

-->

# Distance-based functional classification [Marie]

In this section, we explore whether our geodesic distance estimator has benefits for downstream analysis task. There are many tasks we could consider here such as distance-based nonparametric regression and distanced-based functional clustering, but we will focus on distance-based functional classification. It must be noted that while curve alignment, also known as curve registration, is necessarily performed as a preprocessing technique prior to clustering and classification, our geodesic distance estimator allows one to forsake this step. 

For simplicity, assume the task is binary classification. Associated to each functional object $x$ is a binary $y$ indicating class membership.
Consider the classifier proposed in Ferraty and Vieu (2003,2006) which is a functional version of the Nadaraya-Watson kernel estimator of class membership probabilities:
$$
\hat p(y = 0 | x) \frac{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] 1(y_i = 0) }{ \sum_{i=1}^n K[h^{-1} d(x,x_i)] }
$$
We shall compare our method to using $L_2$ distance, possibly weighted, and with curve registration already accomplished. Describe alternative methods in detail. 

The bandwidth in the classifier should be tuned individually for each method. Also we might need to tune MDS dimension $s$ since in real data, the dimension of the manifold might be much higher than encountered in the simulation scenarios where it never goes above 2. 

Datasets used by functional classification papers
\begin{itemize}
\item Wheat, rainfall and phoneme in Aurore's paper "Achieving near-perfect classification for functional data"
\item Berkeley growth curves in [@ChenReiss2014].
\item Tecator and phoneme in [@Galeano2015] Mahalanobis technometrics paper.
\item yeast cell cycle gene expression (can't find this publicly) in [@LengMuller2005] "Classification using functional data analysis for temporal gene expression data"
\end{itemize}

Datasets used in functional manifold papers
\begin{itemize}
\item Berkeley growth, yeast cell cycle gene expression (can't find this publicly) in [@ChenMuller2012]
\item Tecator in [@LinYao2017] contamination paper
\item Berkeley growth, gait cycle in [@Dimeglio2014] robust isomap paper
\end{itemize}


<!-- ## Distance-based functional clustering -->

<!-- Describe the $k$-medoids clustering method. -->

<!-- ### Simulated data -->

<!-- ###  Real data -->

# References